HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-06 submission.Rmd
increasing caption size

## Introduction
Deciding between man and zone coverage is one of the most critical strategic decisions a defensive coordinator must take before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers coverage predictions during live broadcasts (see a snapshot of the 2024 Week 12 match between the Pittsburgh Steelers and Cleveland Browns). However, in these pre-snap motion does not seem to play an accentuated role (see [Amazon](https://aws.amazon.com/de/blogs/machine-learning/identifying-defense-coverage-schemes-in-nfls-next-gen-stats/)), although it is a crucial element of modern offensive strategies and it is often used to detect the hidden defensive scheme.

@@

#### 1. Pre-motion models

@@

#### 2. Naive post-motion models

@@

#### 3. HMM post-motion models

##### 3.1 Hidden Markov model

@@

##### 3.2 Enhanced post-motion model
We re-train the post-motion model to predict whether the defense plays a man- or zone coverage scheme, but, in this step, we incorporate results from the HMM analysis as further features. However, a caveat of the aforemententioned state probabilities is that they constitute as high-dimensional time-series data which makes them hard to use as features in the previously described models. An outlet of this problem is to employ suitable summary statistics. To overcome this, we first calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$, where $T_n$ is the length of the $n$-th play, and count the number of state switches, i.e. the number of times for which a defender's most likely guarded offensive players differs between two consecutive time points $s$ and $t$, $s>t$. From this, we can calculate some simple, summary statistics: 1) the sum of state switches, 2) the average number of state switches and 3) the number of defenders that switch offensive players during a play. Fourth, we use the aforementioned state decodings to calculate a more elaborated statistic, the mean entropy across defenders $j = 1, \ldots, 5$ in each offensive play $n = 1,\ldots, N$, i.e.:

@@

#### Model and feature evaluation

@@

#### Team analyses

@@

#### Feature engineering

@@

#### Model features and comparison

@@

#### Hidden Markov model

>>>>

## Introduction

Deciding between man and zone coverage is one of the most critical strategic decisions a defensive coordinator must take before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers coverage predictions during live broadcasts (see a snapshot of the 2024 Week 12 match between the Pittsburgh Steelers and Cleveland Browns). However, in these pre-snap motion does not seem to play an accentuated role (see [Amazon](https://aws.amazon.com/de/blogs/machine-learning/identifying-defense-coverage-schemes-in-nfls-next-gen-stats/)), although it is a crucial element of modern offensive strategies and it is often used to detect the hidden defensive scheme.

@@

### 1. Pre-motion models

@@

### 2. Naive post-motion models

@@

### 3. HMM post-motion models

#### 3.1 Hidden Markov model

@@

#### 3.2 Enhanced post-motion model

We re-train the post-motion model to predict whether the defense plays a man- or zone coverage scheme, but, in this step, we incorporate results from the HMM analysis as further features. However, a caveat of the aforemententioned state probabilities is that they constitute as high-dimensional time-series data which makes them hard to use as features in the previously described models. An outlet of this problem is to employ suitable summary statistics. To overcome this, we first calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$, where $T_n$ is the length of the $n$-th play, and count the number of state switches, i.e. the number of times for which a defender's most likely guarded offensive players differs between two consecutive time points $s$ and $t$, $s>t$. From this, we can calculate some simple, summary statistics: 1) the sum of state switches, 2) the average number of state switches and 3) the number of defenders that switch offensive players during a play. Fourth, we use the aforementioned state decodings to calculate a more elaborated statistic, the mean entropy across defenders $j = 1, \ldots, 5$ in each offensive play $n = 1,\ldots, N$, i.e.:

@@

### Model and feature evaluation

@@

### Team analyses

@@

### Feature engineering

@@

### Model features and comparison

@@

### Hidden Markov model
